Generalized Bradley-Terry models

### 2025-03-20 inst/jss_hyper3_real.Rnw
typo
JSS house style

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[article]{jss}
\usepackage{booktabs}
\usepackage{units}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{wrapfig}

%  Process with:
%  
%  Rscript -e 'library("knitr") ; knit("jss_hyper3_real.Rnw")' && pdflatex jss_hyper3_real && open jss_hyper3_real.pdf
%
%  (also the Makefile has an entry for this)

\author{Robin K. S. Hankin\\University of Stirling}
\title{Generalized Bradley-Terry models}
\Plainauthor{Robin K. S. Hankin} %% comma-separated
\Plaintitle{Generalized Bradley-Terry models}

%% an abstract and keywords
\Abstract{

  A generalization of the Bradley-Terry model is presented in which
  strengths may be modified using a multiplicative factor $\lambda$;
  Generally, ordinary Bradley-Terry is recovered if $\lambda=1$.  This
  approach offers multiple advantages over other generalizations such
  as reified Bradley-Terry.  It becomes possible to represent
  non-independence of successive observations in Plackett-Luce
  likelihood functions which allows a parametrization of the ``red
  bus, blue bus'' problem in preference analysis.  I present examples
  drawn from the fields of food preference and the e-sport {\em
    Counterstrike}.  }

\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Robin K. S. Hankin\\
  University of Stirling\\
  E-mail: \email{hankin.robin@gmail.com}
}

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}
\newcommand{\home}{\ensuremath{\mathfrak{H}}}
\newcommand{\oo}{\vphantom{W^{W^{W^W}}}}

@@

<<setup, cache=FALSE,echo=FALSE>>=
source("loadlibrary.R",echo=FALSE)  # to avoid stupid messages
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(prompt = "R> ")
@

\section{Introduction}

\begin{wrapfigure}{r}{0.1\textwidth}
    \centering
    \includegraphics[width=0.1\textwidth]{hyper2.png}\\
\end{wrapfigure}

@@

Further, one might follow~\cite{hankin2017_rnw} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function

\begin{equation}\label{like1234}
  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
  \frac{
    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
    \left(p_3 + \mu p_4\right)^{w_{34}}
    }{
    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
    }
\end{equation}

where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the {\tt hyper2} R
package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors with new \proglang{S3} class {\tt hyper3} and
associated methods.

The most general likelihood function would be
proportional to

\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
  \end{equation}

where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}

\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}

where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.

\subsection{The package in use}
Package idiom for creating a likelihood function for this dataset is
straightforward.  Considering first only Rangers-Celtic matches:

<<defineoldfirm, cache=TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda            )] %<>%      add(81     )
    H[c(                  Celtic = 1)] %<>%      add(     42)
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@

Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:

<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@

We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:

<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
lam <- seq(from=0.9, to=3, len=23)
logLike <- sapply(lam,f)
@

Thus {\tt f()} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

Then using this value we may estimate the Bradley-Terry strengths of
Rangers and Celtic:

<<mleoldfirm,cache=TRUE>>=
maxp(oldfirm(lambda_max))
@

Compare with the theoretical value of $\widehat{p_{\mbox{\tiny
      C}}}=\frac{\sqrt{42\cdot 76}}{\sqrt{81\cdot 46}+\sqrt{42\cdot
    76}}\simeq 0.4807$, $\widehat{p_{\mbox{\tiny R}}}
=\frac{\sqrt{81\cdot 46}}{\sqrt{81\cdot 46}+\sqrt{42\cdot 76}}\simeq
0.5193$, and $\widehat{\lambda}=\frac{\sqrt{76\cdot 81}}{\sqrt{46\cdot
    42}}\simeq 1.785$.

@@

\begin{table}
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}

We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke package function
\code{home\_draw\_away3()} converts data in the form of
Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
Table~\ref{windrawloss}; it takes numerical values for the home
advantage $\lambda$ and proclivity to draw $D$:

<<eg_RCLF3_lf, cache=TRUE>>=
home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
@

Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.

<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
@

Thus \code{f()} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
\code{optim()}:

<<findmaxlambdaD, cache=TRUE>>=
maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
maxv
@

Then

<<showevaluate, cache=TRUE>>=
maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
@

Such estimates have rich predictive power.  For example, if Rangers
were to play Celtic with both teams away [thus nullifying the home
  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
probability of win/draw/loss for Rangers would be
$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
estimated at $(0.39;0.27;0.34)$.

<<makecontourdata, echo=FALSE, cache=TRUE>>=
n <- 20
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

@@

<<testlambdazero, cache=TRUE, include=FALSE>>=
jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
@

Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
  jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975} likelihood function for
this observation:

\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}

Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \code{cheering()} which
implements this functionality.

@@

\begin{figure}[htbp]
  \begin{center}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw[thick] (0,0) -- (2, 3) -- (4.8, 3);
\draw[thick] (0,0) -- (2, 2) -- (4.8, 2);
\draw[thick] (0,0) -- (2, 1) -- (4.8, 1);
\draw[thick] (0,0) -- (2, 0) -- (4.8, 0);
\draw[thick] (0,0) -- (2,-1) -- (4.8,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};

>>>>

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

%% another package (only for this demo article)
\usepackage{framed}
\usepackage{booktabs}
\usepackage{units}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{wrapfig}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("hyper2")
library("magrittr")
howmany <- 23  # set to a smaller value to save time
@

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Robin K. S. Hankin~\orcidlink{0000-0001-5982-0415}\\University of Stirling}
\Plainauthor{Robin K. S Hankin}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title

\title{Generalized Bradley-Terry models}

%% - \Abstract{} almost as usual
\Abstract{
  A generalization of the Bradley-Terry model is presented in which
  strengths may be modified using a multiplicative factor $\lambda$;
  Generally, ordinary Bradley-Terry is recovered if $\lambda=1$.  This
  approach offers multiple advantages over other generalizations such
  as reified Bradley-Terry.  It becomes possible to represent
  non-independence of successive observations in Plackett-Luce
  likelihood functions which allows a parametrization of the ``red
  bus, blue bus'' problem in preference analysis.  I present examples
  drawn from the fields of food preference and the esport {\em
    Counterstrike}.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{JSS, style guide, comma-separated, not capitalized, \proglang{R}}
\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}
\Keywords{Plackett-Luce, Bradley-Terry, \proglang{R}}
\Plainkeywords{Plackett-Luce, Bradley-Terry, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Robin K. S. Hankin\\
  University of Stirling\\
  E-mail: \email{hankin.robin@gmail.com}
}

@@

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section{Introduction}
\begin{wrapfigure}{r}{0.1\textwidth}
    \centering
    \includegraphics[width=0.1\textwidth]{hyper2.png}\\
\end{wrapfigure}

@@

Further, one might follow~\cite{hankin2017_rnw} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function
%
\begin{equation}\label{like1234}
  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
  \frac{
    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
    \left(p_3 + \mu p_4\right)^{w_{34}}
    }{
    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
    }
\end{equation}
%
where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the \pkg{hyper2} R
package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors with new \proglang{S3} \class{hyper3} and
associated methods.

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR}.

\subsection{The package in use}
Package idiom for creating a likelihood function for this dataset is
straightforward.  Considering first only Rangers-Celtic matches:
%
<<defineoldfirm, cache=TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda            )] %<>%      add(81     )
    H[c(                  Celtic = 1)] %<>%      add(     42)
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
lam <- seq(from=0.9, to=3, len=howmany)
logLike <- sapply(lam,f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

Then using this value we may estimate the Bradley-Terry strengths of
Rangers and Celtic:
%
<<mleoldfirm,cache=TRUE>>=
maxp(oldfirm(lambda_max))
@
%
Compare with the theoretical value of $\widehat{p_{\mbox{\tiny
      C}}}=\frac{\sqrt{42\cdot 76}}{\sqrt{81\cdot 46}+\sqrt{42\cdot
    76}}\simeq 0.4807$, $\widehat{p_{\mbox{\tiny R}}}
=\frac{\sqrt{81\cdot 46}}{\sqrt{81\cdot 46}+\sqrt{42\cdot 76}}\simeq
0.5193$, and $\widehat{\lambda}=\frac{\sqrt{76\cdot 81}}{\sqrt{46\cdot
    42}}\simeq 1.785$.

@@

\begin{table}[t!]
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}
%
We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke package function
\code{home\_draw\_away3()} converts data in the form of
Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
Table~\ref{windrawloss}; it takes numerical values for the home
advantage $\lambda$ and proclivity to draw $D$:
%
<<eg_RCLF3_lf, cache=TRUE>>=
home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
@
%
Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.
%
<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
@
%
Thus \code{f()} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
\code{optim()}:
%
<<findmaxlambdaD, cache=TRUE>>=
maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
maxv
@
%
Then
%
<<showevaluate, cache=TRUE>>=
maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
@
%
Such estimates have rich predictive power.  For example, if Rangers
were to play Celtic with both teams away [thus nullifying the home
  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
probability of win/draw/loss for Rangers would be
$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
estimated at $(0.39;0.27;0.34)$.
%
<<makecontourdata, echo=FALSE, cache=TRUE>>=
n <- howmany
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

@@

<<testlambdazero, cache=TRUE, include=FALSE>>=
jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$-value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \code{cheering()} which
implements this functionality.

@@

\begin{figure}[t!]
  \begin{center}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw[thick] (0,0) -- (2, 3) -- (4.8, 3);
\draw[thick] (0,0) -- (2, 2) -- (4.8, 2);
\draw[thick] (0,0) -- (2, 1) -- (4.8, 1);
\draw[thick] (0,0) -- (2, 0) -- (4.8, 0);
\draw[thick] (0,0) -- (2,-1) -- (4.8,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};
